\documentclass{article}

\usepackage[letterpaper,margin=0.5in]{geometry}
\usepackage{parskip}

\usepackage[backend=bibtex,style=ieee]{biblatex}
\bibliography{written_project_proposal}

\title{Towards a Verified Bzip2: A Functional Specification for Burrows-Wheeler}
\author{Jake Waksbaum '20\\Advisor: Andrew Appel}
\date{}

\begin{document}
\maketitle

\section{Motivation and Goal}
File compression software is often used to reduce the amount of disk
space occupied by archived files and to more efficiently transfer
data. Many types of file compression software, such as bzip2, 7zip,
unzip, have been found vulnerable to security exploits that can be
used in denial-of-service attacks or to enable code execution. Often
these vulnerabilities are the result of programmer errors related to
the management of low-level resources: buffer-overflow,
use-after-free, double-free, etc. These types of errors are
well-addressed using formal verification methods that demand proof
that low-level resources are correctly managed.

The goal of my project is to begin the work of producing a formally
verified implementation of the popular and open-source bzip2
compression algorithm. A formally verified compression and
decompression program won't suffer from the types of security exploits
that are so common in existing implementations.

\section{Problem Background and Related Work}
Coq is a theorem proving assistant that allows programmers to write
machine-checked proofs: proofs are verified by Coq to be sound. It
also includes a functional programming language called Gallina that
can be used to produce functional specifications\cite{7536361} that
are useful for describing the desired behavior of some C program.

The CompCert
compiler\cite{Stewart:2015:CC:2676726.2676985,leroy:inria-00000963} is
an optimizing C compiler, that has been formally verified to preserve
semantics: this means that the semantics of the executable code with
resepect to the semantics of the assembly language is the same as the
semantics of the source code with respect to the semantics of the
source langage, C. This allows us to have confidence that the
properties that we prove about our C programs also hold on the
compiled executables.

The Verified Software
Toolchain\cite{Appel:2011:VST:1987211.1987212,Appel:2014:PLC:2670099}
includes also a program logic for C called Verifiable C that is proved
sound with respect to the operational semantics of Compcert C. This
gives us tools in Coq to reason about our C programs to prove
properties such as safety. It includes
VST-Floyd\cite{cao2018vst-floyd:}, which helps automate some parts of
the the process of building functional correctness proofs of C
programs.

These tools were used to formalize the functional specifications of
the mbedTLS HMAC-DRBG, to prove that the specification truly produces
pseuodrandom output, and then to produce an executable for which that
result also holds\cite{Ye:2017:VCS:3133956.3133974}.

\section{Approach}
I plan to work towards a formally verified implementation of bzip2 by
implementing functional specifications for the core compression and
decompression passes that make up the bzip2 stack: run-length
encoding, Burrows-Wheeler, Huffman encoding, etc. This means writing a
functional program in Gallina that implements those algorithms, and
proving useful properties about those algorithms such as that the
decoder inverts the encoder. I will focus on the functional
specification of the algorithims as opposed to the verification of C
implementations.

\section{Plan}
I will start by working on a functional specification for run-length
encoding, as it is simple and the first phase of compression. Then, I
will work on a functional specification for Burrows-Wheeler. Depending
on how long that takes, I can move on to the rest of the compression
stack, implementing and proving theorems about as much of it as I can.

\section{Evaluation}
My main goal is to produce a functional specification for
Burrows-Wheeler and a proof that the decoder inverts the encoder. A
stretch goal would be to verify the entire bzip2 compression stack,
and an even bigger stretch goal would be to produce an entire Coq
program that can operate on real bzip2 zip files.

\printbibliography{}

\end{document}
